---
title: "cone_nmf_f_3"
author: "zihao12"
date: "2021-06-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
I want to confirm my observation that cone NMF (F norm) solution converges to where $B, W^t$ are the same, on realistic data. Here I fit it on a document [dataset](https://zihao12.github.io/ebpmf_data_analysis/kos_K20_ebpmf.alpha_v0.3.8.html) with $(n, p) = (3430, 6906)$

```{r warning=FALSE, message=FALSE}
rm(list = ls())
source("code/cone_NMF_l2.R")
source("code/smallsim2_functions.R")
source("code/misc.R")
source("code/util.R")
set.seed(123)
```


# Fit for 100 iterations
Note it has not converged it. 
```{r}
fitted = readRDS("output/fitted_kos_coneNMF_F_K20.Rds")
plot(fitted$loss)
fitted$runtime
```

Indeed $B, W^t$ are basically identical.   
```{r fig.height=15, fig.width=20}
k = ncol(fitted$B)
par(mfrow = c(5, 4))
for(i in 1:k){
  plot(fitted$B[,i], fitted$W[i,], xlab = "bk",ylab = "wk", cex.lab = 1)
  abline(a = 0, b = 1, col = "blue")
  
}
```


<!-- # Fit for 1000 iterations -->
<!-- ```{r} -->
<!-- fitted = readRDS("output/fitted_kos_coneNMF_F_K20_iter1000.Rds") -->
<!-- plot(fitted$loss) -->
<!-- fitted$runtime -->
<!-- ``` -->

<!-- Indeed $B, W^t$ are basically identical. Will show the results after 1000 iterations when ready.    -->
<!-- ```{r fig.height=15, fig.width=20} -->
<!-- k = ncol(fitted$B) -->
<!-- par(mfrow = c(5, 4)) -->
<!-- for(i in 1:k){ -->
<!--   plot(fitted$B[,i], fitted$W[i,], xlab = "bk",ylab = "wk", cex.lab = 1) -->
<!--   abline(a = 0, b = 1, col = "blue") -->
<!-- } -->
<!-- ``` -->

##  compare with regular NMF
Initialized from coneNMF and ran only 20 iterations till converged (`rel.tol = 1e-04`)
```{r}
fitted_nmf = readRDS("output/fitted_kos_NMF_F_K20.Rds")
n= ncol(fitted_nmf$H)
p= nrow(fitted_nmf$W)


## mse: NMF vs coneNMF
c(min(fitted_nmf$target.loss), min(fitted$loss)/(n*p*2))
```


Compare L
```{r fig.height=15, fig.width=20}
par(mfrow = c(5, 4))
for(i in 1:k){
  plot(fitted$W[i,], fitted_nmf$H[i,],xlab = "Lk (coneNMF)",ylab = "Lk (NMF)", cex.lab = 1)
  abline(a = 0, b = 1, col = "blue")
}
```


Compare F
```{r fig.height=15, fig.width=20}
par(mfrow = c(5, 4))
for(i in 1:k){
  plot(fitted$A[,i], fitted_nmf$W[,i],xlab = "Fk (coneNMF)",ylab = "Fk (NMF)", cex.lab = 1)
  abline(a = 0, b = 1, col = "blue")
}
```


