---
title: "ebpmf.alpha_v0.4.2_show_topics"
author: "zihao12"
date: "2020-06-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
On [kos](https://archive.ics.uci.edu/ml/datasets/Bag+of+Words) dataset,  I show the topics computed from all methods, which are all initialized with the same $L, F$. \

* `bg`: [ebpmf_bg](https://zihao12.github.io/ebpmf_data_analysis/ebpmf_bg.pdf); pick the words with most weight sin each topic\

* `wbg`: [ebpmf_wbg](https://zihao12.github.io/ebpmf_data_analysis/ebpmf_wbg.pdf); pick the words with most weight sin each topic\

* `pmf`: Poisson NMF; pick the words with most weight sin each topic\

* `pmf2binom`: apply `fasTopics::poisson2binom` to PMF results, then use $log(p_1/p_0)$ to rank the importance of words. 

```{r}
library('knitr')
```

## $K = 20$
```{r}
K = 20
top_words_df = read.csv(sprintf("output/uci_BoW/v0.4.2/top_words_K%d_v0.4.2.csv", K))
kable(top_words_df, format = "markdown")
```


## $K = 50$
```{r}
K = 50
top_words_df = read.csv(sprintf("output/uci_BoW/v0.4.2/top_words_K%d_v0.4.2.csv", K))
kable(top_words_df, format = "markdown")
```

## $K = 100$
```{r}
K = 100
top_words_df = read.csv(sprintf("output/uci_BoW/v0.4.2/top_words_K%d_v0.4.2.csv", K))
kable(top_words_df, format = "markdown")
```




